---
title: "NOAA Storm Database - Analysis of event effects"
Synopsis: "max of 10 sentences describing the analysis"
author: "Stefan Buchholz"
date: "Thursday, November 06, 2014"
output: html_document
---

## Data Processing

### Loading Data
Firstly the Data is downloaded from the Coursera link
```{r downloading}
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'

if (file.exists('Storm.Data.csv.bz2') == FALSE) {
    download.file(url, dest='Storm.Data.csv.bz2') }
```

after we downloaded the data we extract the data and read it into a dataframe:

```{r extracting and reading cache=TRUE}
library(R.utils)
bunzip2('Storm.Data.csv.bz2', dest='Storm.Data.csv')
bz <- read.csv('Storm.Data.csv')
```

The columns we need to answer our questions are EVTYPE, FATALITIES, INJURIES,
PROPDMG, CROPDMG.
```{r shrink}
needed <- bz[,c(8,23,24,25,27)]
```

there are no NA values and everything is heavily left skewed.
When we explore the given Storm reasons we can see that we have to clear them up and 
make them more conform by:

1. delete rows where there is a "summary" because that is no interessting data
2. transform them to lower case
3. replace all special characters to blanks
4. strip blank values
5. set new levels

```{r}
#delete everything that looks like a summary line
filter <- grep("summar",needed$EVTYPE, ignore.case = TRUE)
needed <- needed[-filter,]

#set to lower
needed$EVTYPE <- tolower(needed$EVTYPE)

#replace few bad chars
chars <- c("/", "\\", "(", "\)" ,"&")
for(i in 1:length(chars)) {
    needed$EVTYPE <- gsub(chars[i], " ", needed$EVTYPE, fixed=TRUE)
}

#strip out SPACEs
needed$EVTYPE <- gsub(" ", "", needed$EVTYPE, fixed=TRUE)

#set as new levels
needed$EVTYPE <- as.factor(needed$EVTYPE)
table(needed$EVTYPE)
```

## Analysis

Required at least one plot
No more than three plots

Questions to answer:
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

```{r question_1}
needed$health <- needed$FATALITIES + needed$INJURIES

#group by Event type and sum up the health column
library(dplyr)
event_group <- group_by(needed, EVTYPE)
event_group_health <- summarise(event_group,
    health = sum(health))
answer1 <- arrange(event_group_health, desc(health))
```

Now we can plot our answer 1 ... only the first 10 most harmful events will be printed.

```{r print_question1}
answer1_short <- head(answer1, 10)
library(ggplot2)
ggplot(data=answer1_short, aes(x = EVTYPE, y = health)) + 
    geom_bar(stat="identity", position=position_dodge(), colour="black") +
    ggtitle("Plot of the 10 most harmful wheather events in the US") +
    ylab("Number of human fatalities and injuries") +
    xlab("Type of wheather event") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    theme(plot.title = element_text(lineheight=.8, face="bold"))
```



2. Across the United States, which types of events have the greatest economic consequences?

```{r question_2}
needed$economic <- needed$PROPDMG + needed$CROPDMG

#group by Event type and sum up the health column
library(dplyr)
event_group <- group_by(needed, EVTYPE)
event_group_economic <- summarise(event_group,
    economic = sum(economic))
answer2 <- arrange(event_group_economic, desc(economic))
```

now we can plot our answer 1 ... only the first 10 most harmful events will be printed

```{r print_question2}
answer2_short <- head(answer2, 10)
library(ggplot2)
ggplot(data=answer2_short, aes(x = EVTYPE, y = economic)) + 
    geom_bar(stat="identity", position=position_dodge(), colour="black") +
    ggtitle("Plot of the 10 most harmful wheather events in the US") +
    ylab("Number of economic damage") +
    xlab("Type of wheather event") + 
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    theme(plot.title = element_text(lineheight=.8, face="bold"))
```

## Results

As you can see the most harmful wheather event for both, the economy and humanity are tornados. After that the events that make up the most damage in any category are pretty diversified.

