---
title: "NOAA Storm Database - Analysis of event effects"
Synopsis: "max of 10 sentences describing the analysis"
author: "Stefan Buchholz"
date: "Thursday, November 06, 2014"
output: html_document
---

## Data Processing

### Loading Data
Firstly the Data is downloaded from the Coursera link
```{r downloading}
url <- 'https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2'

if (file.exists('Storm.Data.csv.bz2') == FALSE) {
    download.file(url, dest='Storm.Data.csv.bz2') }
```

after we downloaded the data we extract the data and read it into a dataframe:

```{r extracting and reading cache=TRUE}
library(R.utils)
bunzip2('Storm.Data.csv.bz2', dest='Storm.Data.csv')
bz <- read.csv('Storm.Data.csv')
```

The columns we need to answer our questions are EVTYPE, FATALITIES, INJURIES,
PROPDMG, CROPDMG.
```{r shrink}
needed <- bz[,c(8,23,24,25,27)]
```

there are no NA values and everything is heavily left skewed.
When we explore the given Storm reasons we can see that we have to clear them up and 
make them more conform by:

1. strip blank values
2. transform them to lower case
3. delete all special characters

## Analysis

Required at least one plot
No more than three plots

Questions to answer:
1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

2. Across the United States, which types of events have the greatest economic consequences?

## Results
